A Large and Variable Leading Tail of Helium in a Sub-Saturn Undergoing Runaway Inflation

### 2023-07-12 paper/paper1/ms.tex
Final few tiny prose edits from my re-reading

The large size of the gas stream overfills the transit chord, which means both the advancing blue side and receding red side of the rotating star are nearly continuously obscured in this spin-orbit-aligned system \citep{2017AJ....153..211Z}. The planet transits near the stellar equator \citep{2017AJ....153..211Z}, meaning the transit chord covers about 10$\%$ of the projected stellar surface area.  Hypothetically a nearly 100$\%$ opaque optically thick stream of gas covering only the transit chord could reproduce the observed 10$\%$ absorption excess.  Alternatively---and more likely---an optically thin gas stream has some additional vertical extent perpendicular to the plane of orbit, which would act to overfill the entire projected stellar disk.  Assuming the gas has a spatially uniform optical depth of about 0.1 would reproduce the 10\% absorption excess.

@@

When applied to an ensemble of planet systems, both the XUV irradiation and Ohmic dissipation mechanisms expect a void in the planet mass-radius plane. However, the theories make different quantitative predictions for the placement of the dividing lines between stable and unstable populations.  We illustrate these differences in Figure \ref{fig:tf2018_theory}, which shades the mass-radius plane with predictions for the inflation timescale under HAT-P-67b-like conditions.  The XUV irradiation timescale better matches the distribution of planets, with the 0.1 $g\;cm^{-3}$ density contour setting a conspicuous dividing line in density.  The Ohmic dissipation shading predicts short inflation timescales extending into a region of Jupiter-mass planets with radii commonly observed.  The better match of observed exoplanet demographics to the XUV irradiation shading disfavors Ohmic dissipation.

@@

Alternatively, Star Planet Magnetic Interaction (SPMI) could be at play.  In this scenario, the magnetic field of the planet permeates the space between the star and the planet.  Magnetic perturbations propagate via planetary magnetic fields with the Alfv\'en speed.  The planet can interact with the star if the Alfv\'en speed exceeds the stellar wind speed controlling the bulk motion of the intervening medium.  The non-detection of variability in the \ion{Ca}{2} H and K lines and H$\alpha$ lines appears to disfavor this SPMI interpretation since SPMI could be expected to cause variations in these diagnostics.

@@

We have presented a multi-year spectroscopic survey of HAT-P-67~b, a low density, heavily irradiated Saturn-mass (or lower) planet.  We identified a large leading tail, with a much weaker trailing tail, which we found to be consistent with preferential dayside mass loss.  HAT-P-67~b stands out as an outlier in the mass-radius plane, which we examine through the lens of different mechanisms for anomalous heating.  Both XUV irradiation and Ohmic dissipation predict runaway inflation for such inflated hot Saturns, and we quantitatively weigh these two scenarios, finding XUV irrdation as more straightforwardly predictive of the overall demographic population of exoplanets observed to date.

>>>>

The large size of the gas stream overfills the transit chord, which means both the advancing blue side and receding red side of the rotating star are nearly continuously obscured in this spin-orbit-aligned system \citep{2017AJ....153..211Z}. If the planet transits near the stellar equator \citep{2017AJ....153..211Z}, the transit chord covers about 10$\%$ of the projected stellar surface area.  Hypothetically a nearly 100$\%$ opaque optically thick stream of gas covering only the transit chord could reproduce the observed 10$\%$ absorption excess.  Alternatively---and more likely---an optically thin gas stream has some additional vertical extent perpendicular to the plane of orbit, which would act to overfill the entire projected stellar disk.  Assuming the gas has a spatially uniform optical depth of about 0.1 would reproduce the 10\% absorption excess.

@@

When applied to an ensemble of planet systems, both the XUV irradiation and Ohmic dissipation mechanisms expect a void in the planet mass-radius plane. However, the theories make different quantitative predictions for the placement of the dividing lines between stable and unstable populations.  We illustrate these differences in Figure \ref{fig:tf2018_theory}, which shades the mass-radius plane with predictions for the inflation timescale under HAT-P-67b-like conditions.  The XUV irradiation timescale better matches the distribution of planets, with the 0.1g~cm$^{-3}$ density contour setting a conspicuous dividing line in density.  The Ohmic dissipation shading predicts short inflation timescales extending into a region of Jupiter-mass planets with radii commonly observed.  The better match of observed exoplanet demographics to the XUV irradiation shading disfavors Ohmic dissipation.

@@

Alternatively, Star Planet Magnetic Interaction (SPMI) could be at play \citep{2018haex.bookE..25S}.  In this scenario, the magnetic field of the planet permeates the space between the star and the planet.  Magnetic perturbations propagate via planetary magnetic fields with the Alfv\'en speed.  The planet can interact with the star if the Alfv\'en speed exceeds the stellar wind speed controlling the bulk motion of the intervening medium.  The non-detection of variability in the \ion{Ca}{2} H and K lines and H$\alpha$ lines appears to disfavor this SPMI interpretation since SPMI could be expected to cause variations in these diagnostics.

@@

We have presented a multi-year spectroscopic survey of HAT-P-67~b, a low density, heavily irradiated Saturn-mass (or lower) planet.  We identified a large leading tail, with a much weaker trailing tail, which we found to be direct evidence of preferential dayside mass loss.  HAT-P-67~b stands out as an outlier in the mass-radius plane, which we examine through the lens of different mechanisms for anomalous heating.  Both XUV irradiation and Ohmic dissipation predict runaway inflation for such inflated hot Saturns, and we quantitatively weigh these two scenarios, finding XUV irradiation as more straightforwardly predictive of the overall demographic population of exoplanets observed to date.
